---
title: "Cleaning and Description of Analytic dataset"
Author: "Mary Williams"
date: "2024-11-21"
output: github_document
---
# Sarahy's Data Cleaning

This document for cleaning and tidying the data.

```{r, message=FALSE}
library(tidyverse)
library(rvest)
library(sf)
library(purrr)
library(knitr)
library(readxl)
``` 

```{r, message=FALSE}
nineteen_twentyone_county_lyme = read_csv("new_york_county_lyme.csv") %>%  #read in the NYC lyme data
  janitor::clean_names() %>%  #clean the names 
  select(county_name, event_count,percent_rate ) %>%  # select variables of interest 
  rename(NAME = "county_name", 
         lyme_count = "event_count",
         lyme_rate = "percent_rate"
         ) %>%  # renaming so we can have a common variable when merging 
  mutate( data = "lyme") # created a new variable so double check that merging was accurate

print(nineteen_twentyone_county_lyme)

```


Tidy deer tick data 
```{r, message=FALSE}

tick_surveillance = read_csv("./Deer_Tick_Surveillance__Adults__Oct_to_Dec__excluding_Powassan_virus__Beginning_2008_20241116.csv") %>% 
  janitor::clean_names() %>% # clean the names so they follow normal form 
  filter(year %in% 2019:2021 ) %>%  # filtering from the years of interest that are based from the lyme disease data 
  rename(NAME = "county") %>% # renaming for common identifier
  rename(ticks_tested = "total_tested") %>% 
  select(-total_sites_visited, -school_districts_2016_shp, -local_waterfront_revitalization_program_lwrp_communities, -new_york_zip_codes, -counties, -nys_senate_districts, -a_phagocytophilum_percent, -b_microti_percent, -b_miyamotoi_percent) %>%  # selecting and removing variables of interest , 
mutate(data = "ticks")  # adding a variable to identify merging 


print(tick_surveillance)

```

Read in shp file for NY counties 

```{r, message=FALSE}
ny_county =  st_read("./Counties.shp")  # read in the county shapefile 
 
print(ny_county)

```

Merging the data sets 

```{r, message=FALSE}

# merging the excel sheets first 

NY_lyme_tick = full_join(tick_surveillance, nineteen_twentyone_county_lyme, by = "NAME") %>% 
  select(-data.x, -data.y)

print(NY_lyme_tick)
# now merging the shape file with excel file 

NY_lyme_tick_county = NY_lyme_tick %>%
  left_join(ny_county, by = "NAME") %>% 
  janitor::clean_names() 

# didn't delete the other parts from abbrev and later because they are part of the shapefile and if I delete I have to convert it to a dataframe making the shapefile no longer spatial. 


```



```{r}
# merging and reading in tammy csv file with avg temps 

ny_temp = read_csv("./avg_temp_county.csv", skip =1) %>% 
  janitor::clean_names() %>% 
  select(county:dec_22) %>% 
mutate(county = str_replace_all(county, "(?i) county$", "")) %>% #mutated to remove the suffix county so that the data can be merged and have common identifier
  rename(name = "county") %>%  # renaming to have common identifier 
  drop_na(name) # there are two rows with entirely no data so I dropped them 

tick_lyme_weather = NY_lyme_tick_county  %>% # now merging the entire dataset with avg weather 
  left_join(ny_temp, by = "name") %>% 
  janitor::clean_names() 
         
         
         
```

# Saved Final Dataset 

```{r}
write_csv(tick_lyme_weather, "final_data.csv")

tidied = read_csv("final_data.csv")
```



## Attempt Using NOAA data 
```{r,eval= FALSE}
# importing the weather data 

library(dplyr)
library(rnoaa)

# Get a list of all NY station IDs
stations <- ghcnd_stations()
nystationids <-  stations %>% 
  filter(state == "NY") %>% 
  distinct(id)

# Pull the desired weather data for all of these stations
nydat <- meteo_pull_monitors(nystationids$id, 
                             date_min = "2019-01-01", 
                             date_max = "2021-12-31", 
                             var = c("PRCP", "SNOW", "SNWD", "TMAX", "TMIN"))

# Save the resulting data
save(nystationids, nydat, file = "nynoaadat.RData")

```



```{r, eval=FALSE}
# convert from rdata to csv

load("nynoaadat.RData")

ls()  # This will list all the objects in your environment

str(nydat)


noaa_data <- as.data.frame(nydat)


write.csv(data, file = "nydat.csv", row.names = FALSE)

```



```{r, eval=FALSE}

ny_weather = read_csv("./nydat.csv") %>% 
  select(id, date, prcp,tmax) %>% 
  drop_na(tmax) 


ny_weather <- ny_weather %>%
  group_by(id, date) %>%
  summarise(
    prcp = mean(prcp, na.rm = TRUE),
    tmax = mean(tmax, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  pivot_wider(
    names_from = id,
    values_from = c(date, prcp, tmax)
  )

ny_weather <- ny_weather %>%
  unnest(cols = c(date, prcp, tmax))

print(ny_weather)

summary(tick_lyme_weather)
```
# Mary's Data Analysis
```{r}
#load necessary libraries
library(broom)
```

### Relationship between tick density and average temperature per month 
```{r}
density_vs_avg_temp <- lm(tick_population_density ~ jan_19 + feb_19 + mar_19 + apr_19 + may_19 +jun_19 + jul_19 + aug_19 + sep_19 + oct_19 + nov_19 + dec_19 + jan_20 + feb_20 + mar_20 + apr_20 + may_20 + jun_20 + jul_20 + aug_20 + sep_20 + oct_20 + nov_20 + dec_20+ jan_21 + feb_21 + mar_21 + apr_21 + may_21 + jun_21 + jul_21 + aug_21 + sep_21 + oct_21 + nov_21 + dec_21, data = tick_lyme_weather, na.action = na.exclude) #examine the relationship between tick density and average temperature

broom::tidy(density_vs_avg_temp)
```
This output demonstrates that many of the months show non-significant results, therefore meaning a weak or unclear association of each month with tick density. However some months in 2019, like June (p-value= 0.015) and August (p-value = 0.031), which have significant positive outcomes, and July (p-value < 0.005) had a negative outcome. The intercept was -53.0 with a large of SE of 1,473, and a p-value of 0.971. Therefore  
NEED TO LOOK AT THE REST OF THE VARIABLES

### Relationship between lyme_count and average temperatures per month
```{r}
lyme_vs_avg_temp <- lm(lyme_count ~ jan_19 + feb_19 + mar_19 + apr_19 + may_19 + jun_19 + jul_19 + aug_19 + sep_19 + oct_19 + nov_19 + dec_19 + jan_20 + feb_20 + mar_20 + apr_20 + may_20 + jun_20 + jul_20 + aug_20 + sep_20 + oct_20 + nov_20 + dec_20+ jan_21 + feb_21 + mar_21 + apr_21 + may_21 + jun_21 + jul_21 + aug_21 + sep_21 + oct_21 + nov_21 + dec_21, data = tick_lyme_weather)

broom::tidy(lyme_vs_avg_temp)

broom::tidy(lyme_vs_avg_temp) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```
Lyme count has similar findings to tick density, with strong positive relationships in August 2019, and negative relationships in June 2019. The intercept was -17,100 with a (SE: 3592) and a p-value < 0.005. 

### Relationship between lyme_rate and population variables
```{r}
lyme_rate_vs_population <- lm(lyme_rate ~ tick_population_density + total_ticks_collected, data = tick_lyme_weather)
broom::tidy(lyme_rate_vs_population)
```
This linear regression analyzes the relationship between Lyme rate and tick population density and total ticks collected. Intercept estimate was 121 (SE: 16.4), with a test statistic of 7.39 and p-value<0.005. Therefore the intercept is highly signficant, however the predictors are not. This demonstrates a weak association between Lyme rates and the variables. 

The calculated square mileage estimate was -0.0305 (SE: 0.0159), with a statistic of -1.92 and a p-value<0.005. 
 
### Relationship between lyme_rate and area variables
```{r}
lyme_rate_vs_area <- lm(lyme_rate ~ shape_area  + shape_leng + calc_sq_mi, data = tick_lyme_weather)
broom::tidy(lyme_rate_vs_area)
```
Considering that all other variables but the intercept are weak, there does not appear to be an association between lyme rate and area variables (shape area, shape length, and calculated square miles). 

### Trying the relationship between lyme_rate and area variables, including human population density in 2020
```{r}
lyme_rate_vs_area_human <- lm(lyme_rate ~ shape_area  + shape_leng + calc_sq_mi + pop2020, data = tick_lyme_weather)
broom::tidy(lyme_rate_vs_area_human)
```
Human population was found to be negatively associated(-4.14) and significantly association (p-value<0.005) with lyme rate. 

### Relationship between b_burgdorferi_percent and population variables
```{r}
b_burgdorferi_percent_vs_population <- lm(b_burgdorferi_percent ~ lyme_rate + tick_population_density + total_ticks_collected, data = tick_lyme_weather)
broom::tidy(b_burgdorferi_percent_vs_population)
```
