---
title: "Analytic Dataset"
author: "Sarahy Martinez"
date: "2024-11-16"
output: github_document
---

This document for cleaning and tidying the data.

```{r, message=FALSE}
library(tidyverse)
library(rvest)
library(sf)
library(purrr)
library(knitr)
``` 

```{r, message=FALSE}
nineteen_twentyone_county_lyme = read_csv("new_york_county_lyme.csv") %>%  #read in the NYC lyme data
  janitor::clean_names() %>%  #clean the names 
  select(county_name, event_count,percent_rate ) %>%  # select variables of interest 
  rename(NAME = "county_name", 
         lyme_count = "event_count",
         lyme_rate = "percent_rate"
         ) %>%  # renaming so we can have a common variable when merging 
  mutate( data = "lyme") # created a new variable so double check that merging was accurate

print(nineteen_twentyone_county_lyme)

```


Tidy deer tick data 
```{r, message=FALSE}

tick_surveillance = read_csv("./Deer_Tick_Surveillance__Adults__Oct_to_Dec__excluding_Powassan_virus__Beginning_2008_20241116.csv") %>% 
  janitor::clean_names() %>% # clean the names so they follow normal form 
  filter(year %in% 2019:2021 ) %>%  # filtering from the years of interest that are based from the lyme disease data 
  rename(NAME = "county") %>% # renaming for common identifier
  rename(ticks_tested = "total_tested") %>% 
  select(-total_sites_visited, -school_districts_2016_shp, -local_waterfront_revitalization_program_lwrp_communities, -new_york_zip_codes, -counties, -nys_senate_districts, -a_phagocytophilum_percent, -b_microti_percent, -b_miyamotoi_percent) %>%  # selecting and removing variables of interest , 
mutate(data = "ticks")  # adding a variable to identify merging 


print(tick_surveillance)

```

Read in shp file for NY counties 

```{r, message=FALSE}
ny_county =  st_read("./Counties.shp")  # read in the county shapefile 
 
print(ny_county)

```

Merging the data sets 

```{r, message=FALSE}

# merging the excel sheets first 

NY_lyme_tick = full_join(tick_surveillance, nineteen_twentyone_county_lyme, by = "NAME") %>% 
  select(-data.x, -data.y)

print(NY_lyme_tick)
# now merging the shape file with excel file 

NY_lyme_tick_county = NY_lyme_tick %>%
  left_join(ny_county, by = "NAME") %>% 
  janitor::clean_names() 

# didn't delete the other parts from abbrev and later because they are part of the shapefile and if I delete I have to convert it to a dataframe making the shapefile no longer spatial. 

NY_lyme_tick_county %>% 
  knitr::kable()

print(NY_lyme_tick_county)

```



```{r}
# importing the weather data 

library(dplyr)
library(rnoaa)

# Get a list of all NY station IDs
stations <- ghcnd_stations()
nystationids <-  stations %>% 
  filter(state == "NY") %>% 
  distinct(id)

# Pull the desired weather data for all of these stations
nydat <- meteo_pull_monitors(nystationids$id, 
                             date_min = "2019-01-01", 
                             date_max = "2021-12-31", 
                             var = c("PRCP", "SNOW", "SNWD", "TMAX", "TMIN"))

# Save the resulting data
save(nystationids, nydat, file = "nynoaadat.RData")





```




# Description of data
The data set includes `r nrow(NY_lyme_tick_county)` observations and `r ncol(NY_lyme_tick_county)` variables. Of the variables, all are `numeric` except for those that are characters (name, county_centroid, swis, nysp_zone, dos_ll, nyc) and multipolygon (geometry). These contain information about tick surveillance within New York counties, including the number of ticks collected, the number tested for Lyme disease, and the tick population density. The dataset also includes demographic and spatial data of human populations within each county. The information can help identify the high-risk areas and guide public health interventions. 

```{r}
write_csv(NY_lyme_tick_county, "final_data.csv")

tidied = read_csv("final_data.csv")
```


```{r}







```



