---
title: "Exploratory Analysis"
author: "Kaleb J. Frierson"
date: "2024-11-23"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages:
```{r}
library(tidyverse)
```

Copied and pasted `final_data.csv` from data preparation folder into data analysis folder and am reading it into this document as `ticks`. 
```{r}
ticks = read_csv("final_data.csv")
```
I want a count of counties with complete data for each year: 
```{r}
summary_by_year = 
  ticks |> 
  group_by(year) |> 
  summarize(non_missing_counties = sum(!is.na(total_ticks_collected)))

summary_by_year
```

I also want to know which couties have non zero values for all three years of data collection: 
```{r}
all_year_data = 
  ticks |> 
  filter(year %in% c(2019, 2020, 2021)) |>  
  group_by(name) |>                        
  summarize(
    years_with_data = sum(!is.na(total_ticks_collected)),  
    all_years_present = all(year %in% c(2019, 2020, 2021)) 
  ) |> 
  filter(years_with_data == 3 & all_years_present) |>      
  select(name)                                          

all_year_data

```

I think it would be reasonable to restrict the dataset to the 36 counties with data across three years. This would allow us to incorporate a column for tmax for each county in each year and potentially assess any correlation between temperature increases and abundance. 