---
title: "Cleaning and Description of Analytic dataset"
Author: "Mary Williams"
date: "2024-11-21"
output: github_document
---


This document for cleaning and tidying the data.

```{r, message=FALSE}
library(tidyverse)
library(rvest)
library(sf)
library(purrr)
``` 

```{r, message=FALSE}
nineteen_twentyone_county_lyme = read_csv("new_york_county_lyme.csv") %>% 
  janitor::clean_names() %>% 
  select(county_name, event_count,percent_rate ) %>% 
  rename(NAME = "county_name") %>% 
  mutate( data = "lyme")

print(nineteen_twentyone_county_lyme)
```


Tidy deer tick data 
```{r, message=FALSE}

tick_surveillance = read_csv("./Deer_Tick_Surveillance__Adults__Oct_to_Dec__excluding_Powassan_virus__Beginning_2008_20241116.csv") %>% 
  janitor::clean_names() %>% 
  filter(year %in% 2019:2021 ) %>% 
  rename(NAME = "county") %>% 
  select(-total_sites_visited, -school_districts_2016_shp, -local_waterfront_revitalization_program_lwrp_communities, -new_york_zip_codes, -counties, -nys_senate_districts, -a_phagocytophilum_percent, -b_microti_percent, -b_miyamotoi_percent) %>% 
mutate(data = "ticks")


print(tick_surveillance)
```

Read in shp file for NY counties 

```{r, message=FALSE}
ny_county =  st_read("./Counties.shp")  %>%
  select(-DOS_LL, -DOSLL_DATE)

print(ny_county)
```

Merging the data sets 

```{r, message=FALSE}

# merging the excel sheets first 

NY_lyme_tick = full_join(tick_surveillance, nineteen_twentyone_county_lyme, by = "NAME") %>% 
  select(-data.x, -data.y)

print(NY_lyme_tick)
# now merging the shape file with excel file 

NY_lyme_tick_county = NY_lyme_tick %>%
  left_join(ny_county, by = "NAME") %>% 
  janitor::clean_names()
# didn't delete the other parts from abbrev and later because they are part of the shapefile and if I delete I have to convert it to a dataframe making the shapefile no longer spatial. 

print(NY_lyme_tick_county)
summary(NY_lyme_tick_county)
```

# Description of data
The data set includes `r nrow(NY_lyme_tick_county)` observations and `r nrow(NY_lyme_tick_county)` variables. Of the variables, all are `numeric` except for those that are characters (name, county_centroid, swis, nysp_zone, dos_ll, nyc) and multipolygon (geometry). These contain information about tick surveillance within New York counties, including the number of ticks collected, the number tested for Lyme disease, and the tick population density. The dataset also includes demographic and spatial data of human populations within each county. The information can help identify the high-risk areas and guide public health interventions. 


system("rm .DS_Store")

